ODGI: scalable tools for pangenome graphs

### 2021-06-01 manuscript/main.tex
edits

    \abstract{ \textbf{Motivation:} Pangenomes address the
      shortcomings of mainline genomics workflows, in particular with
      regard to reference genome bias and the identifation of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that makes use of an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes at the same time while maintaining fast
        execution runtime and low memory overheads.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graphs
        manipulation and visualisation.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and user manual can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}

@@

    Thanks to advances in sequencing technologies, new genome
    assemblies are produced at a high rate, including the human
    pangenome reference consortium (HPRC) and telomere-to-telemere
    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
    of genomic elements in a given species or clade. Pangenomics thus
    stands in contrast to reference-based genomic approaches which
    relate sequences to a particular consensus model of the
    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
    represented by a graph data structure incorporating sequences as
    nodes and their connections as edges. These nodes can be shared
    for homologs and orthologs.  A `bidirected' graph represents both
    strands of DNA and so-called paths represent contigs that run
    through the nodes. A path can represent a chromosome or part
    thereof.

    A major challenge is writing software that can deal with the size
    of these graphs. vgtools pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    processing large data (see Table~\ref{tab:02}). With \odgi\ we
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI).  that implements is a memory
    efficient graph structure that can be dynamically updated using
    multiple CPUs. This is particularly relevant for highly repetitive
    regions. Segmental duplications, and acrocentric chromosomes lead
    to graphs which present very high node depths which greatly
    increases processing complexity.

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ includes the following tools

    \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, such as
    validating that paths respect the graph's topology.

>>>>

    \abstract{ \textbf{Motivation:} Pangenomes address the
      shortcomings of mainline genomics workflows, in particular
      concerning reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that makes use of an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes at the same time while maintaining fast
        execution runtime and low memory overheads.  \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci},
        exploratory analysis, sorting, removing artifacts, and graph
        manipulation and visualisation.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and user manual can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME}

@@

    Thanks to advances in sequencing technologies, new genome
    assemblies are produced at a high rate, including the human
    pangenome reference consortium (HPRC) and telomere-to-telemere
    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
    of genomic elements in a given species or clade. Pangenomics
    therefore contrasts with reference genome based approaches which
    relate sequences to a particular consensus model of the
    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
    represented by a graph data structure incorporating sequences as
    nodes and their connections as edges. These nodes are shared for
    homologs, paralogs and orthologs.  A `bidirected' graph represents
    both strands of DNA and so-called paths represent contigs that run
    through the nodes. A path can represent an individual chromosome
    or part thereof. Such a graph-based approach is particularly
    relevant for highly repetitive regions. Segmental duplications,
    and acrocentric chromosomes lead to graphs which present very high
    node depths which greatly increases processing complexity.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    large data (see Table~\ref{tab:02}). With \odgi\ we
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in RAM that can be dynamically updated using
    multiple CPUs.

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ includes the following tools:

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, such as
    validating that paths respect the graph's topology.
